---
title: "Extracting estimated and derived parameters"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting estimated and derived parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(phenomix)
library(dplyr)
```

We will start with the original simple model fit to the `fishdist` data,  

```{r}
data("fishdist")
cov_dat <- data.frame(nyear = unique(fishdist$year))
# rescale year -- could also standardize with scale()
cov_dat$nyear <- cov_dat$nyear - min(cov_dat$nyear)
```


```{r}
datalist <- create_data(fishdist,
  min_number = 0,
  variable = "number",
  time = "year",
  date = "doy",
  asymmetric_model = FALSE,
  mu = ~nyear,
  sigma = ~nyear,
  covar_data = cov_dat,
  est_sigma_re = TRUE,
  est_mu_re = TRUE,
  tail_model = "gaussian"
)
```

```{r message=FALSE, warning=FALSE, results='hide'}
set.seed(1)
fitted <- fit(datalist)
```

## Extracting parameters manually

The first option for extracting parameters manually is via the `sdreport` of TMB objects. This includes both estimated and derived quantities. You can see the names of what's available by looking at 

```{r eval=FALSE}
fitted$sdreport$value
```

or perhaps look at these in tabular form,  
```{r}
table(names(fitted$sdreport$value))
```

So if you wanted to extract the mean phenological peak in each time step (named `mu`), you could look at the indices of the names 

```{r}
idx <- which(names(fitted$sdreport$value) == "mu")
```

And then use the respective means and sds corresponding to those indices,

```{r}
m <- fitted$sdreport$value[idx]
s <- fitted$sdreport$sd[idx]
```

## Helper functions

We've also included some helper functions to extract these more quickly yourself. These all take in a fitted model object, and can be called as 

```{r}
m <- extract_means(fitted) # means
s <- extract_sigma(fitted) # sigmas, describing tails
t <- extract_theta(fitted) # theta parameters (scaling)
u <- extract_upper(fitted) # upper quartile
m <- extract_lower(fitted) # lower quartile
```

Or if you want to extact all of the above, you can use

```{r}
e <- extract_all(fitted)
```

## Annual summaries

In addition to the parameters above, we can extract the derived annual totals (predicted across all days 1-365). These are extracted with the following `extract_annual` function. Note that the second parameter controls whether the estimates are in normal or log space. 

```{r}
est_log <- extract_annual(fitted, log = TRUE)
est_normal <- extract_annual(fitted, log = FALSE)
```


