---
title: "Curve fitting with the salmix package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Curve fitting with the salmix package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618
)
```

```{r packages, message=FALSE, warning=TRUE}
library(salmix)
library(dplyr)
library(TMB)
set.seed(1)
```

We'll start by loading the example dataset, based on run timing data for Pacific salmon.

```{r}
glimpse(fishdist)
```

### Creating data for estimation

There's three main functions we'll work with. First, we'll use the `create_data` function, which builds the data and model arguments to be used for fitting. We'll start with the default arguments, before going into detail about what they all mean. The only argument that we've initially changed from the default is using `asymmetric_model = FALSE` to fit the symmetric model.

```{r}
datalist = create_data(fishdist, 
  min_number=0, 
  variable = "number", 
  time="year", 
  date = "doy",
  asymmetric_model = FALSE, 
  est_sigma_trend = TRUE, 
  est_mu_trend = TRUE, 
  est_t_model = FALSE, 
  family = "gaussian")
```

The `min_number` argument represents an optional threshold below which data is ignored. The `variable` argument is a character identifying the name of the response variable in the data frame. Similarly, the `time` and `date` arguments specify the labels of the temporal (e.g. year) and seasonal variables (day of year).

The `salmix` package can fit asymmetric or symmetric models to the distribution data (whether the left side of the curve is the same shape as the right) and defaults to FALSE. The mean and standard deviations that control the distributions are allowed to vary over time (as random effects) but we can also include temporal trends in the mean and standard deviations. These are estimated by default, and controlled with the `est_sigma_trend` and `est_mu_trend` arguments. The trends are log-linear for the standard deviation parameters, but in normal space for the mean parameters. Finally, the tails of the response curves may be estimated via fitting a Gaussian (normal) distribution or a Student-t distribution. By default, the Gaussian tails are estimated, and this parameter is controlled with the `est_t_model` argument.

Last, we can model the observed count data with a number of different distributions, and set this with the `family` argument. Currently supported distributions include the Gaussian, Poisson, and Negative Binomial, and all include a log-link. 

### Fitting the model

Next, we'll use the `fit` function to do maximum likelihood estimation in TMB. Additional arguments can be found in the help file, but the most important one is the list of data created above.

```{r}
fitted = fit(datalist)
```

We don't get any warnings out of the estimation, but let's look at the sdreport in more detail.

```{r}
sdrep = TMB::sdreport(fitted$obj)
sdrep_df = data.frame("par"=names(sdrep$value),
  "value"=sdrep$value, "sd"=sdrep$sd)
head(sdrep_df)
```




### Diagnosing lack of convergence

In addition to the warning message about the Hessian not being positive definite, the NaNs in the `sd` column are a sure sign that things aren't converging. 

Fixes to improve convergence may be to start from a different set of starting values (try passing `inits` into the `fit` function), or placing stricter bounds on convergence statistics. Or it may be that the model isn't a good fit to the data. But we'll try improving things by adjusting the convergence criterion. The `control` argument is a list that is passed into `stats::nlminb` -- we'll only adjust `rel.tol` in this example, but you can tweak other parameters. 

